ASTR480 Final Report:\\ \textit{Asteroids in TESS}

### 2024-10-19 FinalReport/FinalReportDraft.tex
comp Table made

\usepackage[authoryear,comma,round]{natbib}
\usepackage{hyperref,graphicx}
\usepackage[separate-uncertainty=true,multi-part-units=single]{siunitx}

@@

% \usepackage[margin=1.5cm, left=3cm ,twoside]{geometry}

@@

% \title{ASTR480 Final Report:\\
% \textit{Asteroids in TESS}
% }
% \author{Brayden Leicester \\ 
% \textit{61352063} \\
% [1ex]\small{Supervisors: Michele Bannister and Ryan Ridden-Harper}
% }
\begin{document}

@@

An example of the matches found using this \texttt{KDTree} is shown in \autoref{Fig:RADecMatch}.
It is clear that \texttt{TESSELLATE} is not detecting every frame that an asteroid is in.
Detections in \texttt{TESSELLATE} are done on a per pixel basis, and asteroids only cross each pixel for a short number of frames.
The residuals of \autoref{Fig:RADecMatch} are all within a TESS pixel, shown on the right hand axis of each of the lower pannels.

@@

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{./Figures/perDistr.pdf}
  \caption[Period Distribution]{The distribution of rotation periods in \qty{24}{\hour} (\qty{1}{\day}) bins for all the asteroids that passed the quality checks.}
  \label{Fig:PerDistr}
\end{figure}

The distribution of the amplitude of the lightcurves is given in \autoref{Fig:ampDistr}.
The peak of the distrabution at $\frac13$ \unit{\mag} and is right skewed.
The high $\Delta m$ values are from lightcurves of dubious quality. 
Manually checking the analysis, it is clear that these points above $\Delta m =2$ have all been affected by background contamination, and have untrustworth photometry.
Some examples are given in the appendix for asteroids with known periods.

It is unclear how to avoid such contamination.
For some, analysis of the residuals could show a pattern that leads to a period and possible fit of a model, but this is not the case for all of them.
For only one sector worth of data, it was managable for me to preform a manual check on these few outliers
But this will not be the case for all the whole dataset \texttt{TESSELLATE} will analyse.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{./Figures/delMagDistr.pdf}
  \caption[Amplitude Variation Distribution]{The distribution of amplitude variation for all the asteroids that passed the quality checks. 
  The data are binned in steps of $\frac{1}{8}$ \unit{\mag}.
  }
  \label{Fig:ampDistr}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{./Figures/QualPervsAmp1I.pdf}
  \caption[Amplitude of Variation against Rotation Period]{The amplitude of variation in the model from the period analysis against the calculated rotation period.
    This is for all 374 %!CHECK
    objects that pass the quality checks.
  }
  \label{Fig:perVSamp}
\end{figure}

@@

\subsection{Comparison to other work }%\citet{McNeill2023}}

%TODO redo fig 1

@@

\autoref{Fig:FreqDiamUpdate} shows how my rotation periods fit with those in the LCDB of \cite{Warner2009} from \autoref{Fig:FreqVsDiam}.
It is clear that TESS struggles to determine periods for asteroids with a diameter $\lesssim \qty{1}{\kilo \metre}$. 
This is most likely due to these bodies being too dim for the telescope.  
The spin barrier of \citet{Pravec2000} is not violated by any of the periods I have calculated.

269 of the asteroids that pass the quality checks have diameters in the near-Earth object wide-field infrared survey explorer (NEOWISE) V2.0 database \citep{Masiero2011, NEOWISE2019}.
It is only this sub-sample of the 374 asteroids that are plotted in \autoref{Fig:FreqDiamUpdate}.
The diameters in the LCBD are not calculated using NEOWISE data, so there could be discrepancies in those asteroids that I have calculated periods for that are also in the LCDB. 
The NEOWISE data is self-consistent and encompasses more asteroids that the diameters from just the LCDB, so it was preferred for the over plotting of the rotation data from this work.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{./Figures/M23andP20Comp.pdf}
  \caption[Comparison to periods from other work]{Comparison of the rotation periods I have calculated to the publicly available data from \citet{McNeill2023} (left, pink crosses) and \citet{Pal2020} (right, purple pluses). The asteroids that are in all three are highlighted with black rings around the symbols.}
\end{figure}

18 in \citet{McNeill2023}
13 in \citet{Pal2020}

>>>>

\usepackage[authoryear,comma,round]{natbib}
\usepackage{hyperref,graphicx}
\usepackage{datatool,booktabs,pgfplotstable}
\usepackage[separate-uncertainty=true,multi-part-units=single]{siunitx}

@@

@@

\pgfplotsset{compat=1.18}
\renewcommand*\dtldisplaystarttab{\toprule}
\renewcommand*\dtldisplayendtab{\tabularnewline\bottomrule}
\renewcommand*\dtldisplayafterhead{\midrule}
\pgfplotstableset{
    every head row/.style={before row=\toprule, after row=\midrule},
    every last row/.style={after row=\bottomrule}
}

\begin{document}

@@

An example of the matches found using this \texttt{KDTree} is shown in \autoref{Fig:RADecMatch}.
It is clear that \texttt{TESSELLATE} is not detecting every frame that an asteroid is in.
Detections in \texttt{TESSELLATE} are done on a per pixel basis, and asteroids only cross each pixel for a short number of frames.
The residuals of \autoref{Fig:RADecMatch} are all within a TESS pixel, shown on the right hand axis of each of the lower panels.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{./Figures/perDistr.pdf}
  \caption[Period Distribution]{The distribution of rotation periods in \qty{24}{\hour} (\qty{1}{\day}) bins for all the asteroids that passed the quality checks.}
  \label{Fig:PerDistr}
\end{figure}

@@

\begin{figure}[!hb]
  \centering
  \includegraphics[width=\textwidth]{./Figures/delMagDistr.pdf}
  \caption[Amplitude Variation Distribution]{The distribution of amplitude variation for all the asteroids that passed the quality checks.
    The data are binned in steps of $\frac{1}{8}$ \unit{\mag}.
  }
  \label{Fig:ampDistr}
\end{figure}

The distribution of the amplitude of the lightcurves is given in \autoref{Fig:ampDistr}.
The peak of the distribution at $\frac13$ \unit{\mag} and is right skewed.
The high $\Delta m$ values are from lightcurves of dubious quality.
Manually checking the analysis, it is clear that these points above $\Delta m =2$ have all been affected by background contamination, and have untrustworth photometry.
Some examples are given in the appendix for asteroids with known periods.

It is unclear how to avoid such contamination.
For some objects, analysis of the residuals could show a pattern that leads to a period and possible fit of a model, but this is not the case for all of them.
For only one sector worth of data, it was managable for me to preform a manual check on these few outliers
But this will not be the case for all the whole dataset \texttt{TESSELLATE} will analyse.

\begin{figure}[h]
  \centering
  \includegraphics[width = \textwidth]{./Figures/QualPervsAmp1I.pdf}
  \caption[Amplitude of Variation against Rotation Period]{The amplitude of variation in the model from the period analysis against the calculated rotation period.
    This is for all 374 %!CHECK
    objects that pass the quality checks.
  }
  \label{Fig:perVSamp}
\end{figure}

Comparing the rotation period to the $\Delta m$ is insightful and is shown in \autoref{Fig:perVSamp}.
All 374 of the objects that pass the quality checks are plotted here, and no trend is observed.
The variation is quite constant accros the range of rotation periods, although there are less objects at longer periods.

The modes of each of the distributions in \autoref{Fig:PerDistr} and \autoref{Fig:ampDistr} do not fully overlap, as indicated by \autoref{Fig:perVSamp}.
Those asteroids with rotation periods $<\qty{24}{\hour}$ have a range of rotation periods.
Converly, the objects with a $\Delta m \approx 0.3$ span the entire period range.

The position in these axes of \omuamua is shown in \autoref{Fig:perVSamp} by the black caltrop and is highlighted with text.
1I's rotation period of \qty{8.67(0.34)}{\hour} from \citet{Belton2018} and a $\Delta m$ of \qty{2.5}{\mag} from \citet{Meech2017} are used.
There are some asteroids with a larger $\Delta m$ but, as discussed, these are not trustworthy.

@@

\subsection{Comparison to Literature Data}\label{SubSec:LitComp}

@@

\autoref{Fig:FreqDiamUpdate} displays my calculated rotation periods with those in the LCDB of \citet{Warner2009} (\autoref{Fig:FreqVsDiam}).
269 of the asteroids that pass the quality checks have diameters in the near-Earth object wide-field infrared survey explorer (NEOWISE) V2.0 database \citep{Masiero2011, NEOWISE2019}.
It is only this sub-sample of the 374 asteroids that are plotted in \autoref{Fig:FreqDiamUpdate}.
The diameters in the LCBD are not calculated using NEOWISE data, so there could be discrepancies in those asteroids that I have calculated periods for that are also in the LCDB.
The NEOWISE data is self-consistent and encompasses more asteroids that the diameters from just the LCDB, so it was preferred for the over plotting of the rotation data from this work.

It is clear that TESS struggles to determine periods for asteroids with a diameter $\lesssim \qty{1}{\kilo \metre}$.
This is most likely due to these bodies being too dim for the telescope.
Those that are seen must not pass the quality checks I have imposed.

The spin barrier of \citet{Pravec2000} is not violated by any of the periods calculated.
The quality checks to do exclued this, as the spin barrier is at $\sim \qty{2.4}{\hour}$, and I only impose the period must be $>\qty{1}{\hour}$.
This cadence of FFI has the ability to probe the forbidden region, to find more objects that violate the spin barrier.
Unfortunalty, none were found in sector 29.

\begin{figure}[!t]
  \centering
  \includegraphics[width=\textwidth]{./Figures/M23andP20Comp.pdf}
  \caption[Comparison to TESS periods from the Literature]{Comparison of the rotation periods I have calculated to the publicly available data from \citet{McNeill2023} (left, pink crosses) and \citet{Pal2020} (right, purple pluses). The asteroids that are in all three are highlighted with black rings around the symbols.
    A $y=x$ line is plotted to guide the eye.}
  \label{Fig:TESSLit}
\end{figure}

As mentioned in \autoref{SubSec:Asteroid}, this is not the first work done analysing asteroids in TESS data.
This work has 18 objects in common with \citet{McNeill2023} and an overlap of 13 asteroids with the data of \citet{Pal2020}, shown in \autoref{Fig:TESSLit}.
There is agreement in the derivered periods for some of the objects in both datasets, as seen by the points on or near the $y=x$ line in each subplot.
These datasets are in broad, but not complete, agreement for the asteroids they both measure \citep[see the comparison in ][]{McNeill2023}, so it is not surpising that this work is not in complete agreement.

It does not appear that my calculated periods are biases towards longer or shorter values.
Both in of these comparison datasets, and in the check against the LCDB in \autoref{Fig:LCDBcomp}, there are periods on either extreme.
Sometimes the periods found here are much shorter than those found in the external data, and othertimes much longer, but this is in about equal proportions.
The sample size is really too small to make any statistical statments, once again more sectors need to be analysed.

There are 8 objects that appear in all three datasets.
Two of them lie almost directly on top of each other in the \citeauthor{Pal2020} data on the right panels\autoref{Fig:TESSLit}, but there is a difference on the left hand side in the \citeauthor{McNeill2023} data.
This intersection is shown again in \autoref{Tab:AllPers}, where it is more clear how well some of these data line up.
Both (13364) 1998 UK20 and (16151) 1999 XF230 have the same rotation period measured in all three analysies.
My period for (36240) 1999 VN44 is two orders of magnitude larger than either of the two previos studies, so there must have been an issue with the data for this object in sector 29.

I seem to calculate similar periods to \citeauthor{Pal2020} more frequently than with \citeauthor{McNeill2023}, but again this is too small of a sample size.
I have only analysesed a sector of data, they both measure periods for the same 13 sectors. %!Check
When more data has been run through both \texttt{TESSELLATE} and my pipelines, then the number of intersecting objects will increase.

\begin{table}[!t]
  \centering
  \caption[TESS Literature Comparison Table]{The 8 asteroids have rotation periods in this work rounded to two decimal places, as well as \citet{Pal2020} and \citet{McNeill2023}. All periods are measured in hours (\unit{\hour})} \label{Tab:AllPers}
  \pgfplotstabletypeset[
    col sep=&,
    string type,
    columns/a/.style={column name=(MPC designation) Asteroid Name},
    columns/b/.style={column name=This work},
    columns/c/.style={column name=\citeauthor{McNeill2023}},
    columns/d/.style={column name=\citeauthor{Pal2020}}
  ]{tabDat.dat}
\end{table}
